### 2020-08-05 ART_Dada_Piech/Dada-PU.tex
P. Polak correctons added

\begin{artengenv}{Robert Piechowicz, Jerzy Dadaczyński}
	{Maxime Bôcher's concept of complementary philosophy of mathematics}
	{Maxime Bôcher's concept of complementary philosophy of mathematics}
	{Maxime Bôcher's concept of complementary philosophy\\of mathematics}
	{Pontifical University of John Paul II in Krakow}
	{The main purpose of the present paper is to demonstrate that as early as 1904 pre-eminent American mathematician Maxime Bôcher was an adherent to the presently relevant argument of reasonableness, or even necessity of parallel development of two philosophical methods of reflection on mathematics, so that its essence could be more fully comprehended. The goal of the research gives rise to the question: what two types of philosophical deliberation on mathematics were proposed by Bôcher?}
	{philosophy, mathematics, Maxim Bôcher, metamathematics.}

\section{Introduction}
The main purpose of the present paper is to demonstrate that as early as 1904 pre-eminent American mathematician Maxime Bôcher\footnote{Maxime Bôcher (1867-1918) was a son of an American scientist of French origin. In the years 1883-1888 he studied, inter alia, mathematics at Harvard. In the years 1888-1891, in Göttingen he was a student of Klein's, under the direction of whom he wrote and defended his doctoral dissertation. After his return to the USA he was an assistant professor, and as of 1904 a professor of mathematics at Harvard. In the years 1909-1910 he served as president of the American Mathematical Society. Bôcher specialised chiefly in mathematical analysis, and his most famous paper is \textit{Introduction to Higher Algebra}
%\label{ref:RNDc7ZxUmWWTi}(Bôcher, 1907).
\parencite[][]{bocher_introduction_1907}.
To honour the memory of this mathematician, who died prematurely, Bôcher Memorial Prize was endowed; it is awarded at intervals of several years for achievements in the field of mathematical analysis in the period of the six years leading up to the prize-giving year. One of the prize-winners was von Neumann (in the 1930s).} was an adherent to the presently relevant argument of reasonableness, or even necessity of parallel development of two philosophical -- complementary -- methods of reflection on mathematics, so that its essence could be more fully comprehended.

Thus outlined, the main goal of the research gives rise to the question: what two -- complementary -- types of philosophical deliberation on mathematics were proposed by Bôcher? They were ``adumbrated'' in the text entitled \textit{The Fundamental Conceptions and Methods of Mathematics}
%\label{ref:RNDNx8Z5TUKyi}(Bôcher, 1904).
\parencite[][]{bocher_fundamental_1904}.

In the first, principal part of his text -- which has already been analysed in

@@

\parencite[][]{dadaczynski_tendencje_2015} -- Bôcher made a direct reference to logicism, proposing its corrections, so that the reductionist programme could be appropriately realised.\footnote{Bôcher was doubtful as to the possibility of realising the programme of logicism the way it had been delineated by Frege and Russell. His objection was that logic -- as viewed by the originators of logicism -- was too ``weak'' to ``produce'' an infinite set of natural numbers. Bôcher's predictions proved right while work on \textit{Principia Mathematica} was underway: it was necessary to ``add'' some form of an axiom of infinity -- in the ontological version providing for the existence of an infinitude of individuals -- to obtain natural numbers (of any type). Such an axiom did not ``fit within'' Frege's logic or Russell's ``starting'' logic. The American mathematician formulated his objections as follows: ``Nevertheless, in view of the fact that the system of finite positive integers is necessary in almost all branches of mathematics (we cannot speak of a triangle or a hexagon without having the numbers three and six at our disposal), it seems extremely desirable that the system of logical principles which we lay at the foundation of all mathematics be assumed, if possible, broad enough so that the existence of positive integers -- at least finite integers -- follows from it; and there seems little doubt that this can be done in a satisfactory manner. When this has been done we shall perhaps be able to regard, with Russell, pure mathematics as consisting exclusively of deductions by ‘logical principles from logical principles'' 

@@

} He made an \textit{explicite} demand for future demonstration of non-contradiction of all mathematics liberated -- and such liberation he anticipated -- of known antinomies, and not just ``partial'' demands for proof of non-contradiction of the arithmetic of real numbers 

@@

In a sense -- which was demonstrated in the paper 

@@

\parencite[][]{dadaczynski_tendencje_2015}
 -- the first part of Bôcher's text contains, inter alia, an ``outline,'' though an unordered one, of the 1904 state of philosophy of mathematics confederated with the study of its foundations.\footnote{The passage below serves to bring back the conclusions from the paper 

@@

 In 1904 logicism was already a well-developed direction in the study of the foundations of mathematics. Bôcher could see its intrinsic difficulties concerned with the establishment of antinomies, but he was sure that corrections of the logic underlying the foundations of mathematics would make it possible to remove those difficulties. Hence, in a sense, he ``announced'' the ``second'' logicism -- the one of Russell's. Two points excerpted from the 1904 text foreshadowed Hilbert's formalism of the 1920s. They are, firstly, methodical nominalism, and, secondly, posing a question whether it is possible to establish -- having removed known antinomies -- a global non-contradiction of mathematics. The latter issue was Bôcher's clear reference to Hilbert's second problem (non-contradiction of the arithmetic of real numbers) and Hilbert's first attempts at proving the non-contradiction of the arithmetic of natural numbers. The positive solution to the problem of the global non-contradiction of mathematics -- which was emphasised by, inter alia, Bôcher -- later on became the goal of the formalist research programme, the attainment of which being the rationale for the emergence of metamathematics. Methodological nominalism became Hilbert's answer to the question about semantics of the fully axiomatised and formalised classical mathematics. Bôcher's text of 1904 negatively treats Kantianism as the foundation on which to construct mathematics. Nevertheless, it is a source that confirms that Kantianism was still relevant in many mathematical milieus in the early 20\textsuperscript{th} century. That was the ground on which, several years after the publication of Bôcher's text and following the adoption of appropriate corrections and modifications, intuitionism emerged.} Noteworthily, Bôcher termed his own way of deliberating over mathematics, which he presented in the first, principal part of his work, ``discussion [...] of the so-called foundations'' 

@@

 which focuses on the deductive aspect of mathematics. His -- to use his own term -- ``discussion [...] of the so-called foundations'' of mathematics, in which he develops some of his own threads, inter alia, attempting to specify what mathematics is, is a kind of reflection markedly determined by the way (and ``contents'') of Frege's, Russell's and Hilbert's deliberation on mathematics.

Since the first way of deliberation on mathematics, which was proposed by Bôcher, has already been subjected to analysis and identified, demonstrating that his metaphilosophy contained a demand for inclusion of the two -- complementary -- types of philosophy of mathematics becomes reduced to showing that Bôcher was outlining an alternative -- though naturally not in the sense of an exclusive alternative -- concept of philosophical reflection on mathematics, which in his opinion was to complement philosophy confederated with the study of the foundations of mathematics.

With a view to the attainment of the above-indicated research objective, in the first place, presentation will be made of the main outlines of \textit{quasi}-empiricism -- philosophy of mathematics initiated with Lakatos' publications of the first half of the 1960s. The fundamental imperatives of \textit{quasi}-empiricism shall serve here as points of reference for analysis and identification of the content of the alternative variant of philosophy of mathematics outlined by Bôcher.

This place calls for a crucial remark. Originated with Lakatos' texts, \textit{quasi}-empiricism strongly emphasised some threads of philosophy of mathematics which, although constantly present in deliberation on mathematics, for more than half a century were ``suppressed'' by study concerned with foundations (logicism, formalism and intuitionism). Mention will be made here of the prehistory of \textit{quasi}-empiricism, and one of partial propositions will be to show that Bôcher should be reckoned among its representatives. The choice of \textit{quasi}-empiricism as a point of reference for the study of an alternative thread of Bôcher's philosophy of mathematics may seem to be an ahistorical approach exactly because it does not take account of the above-mentioned chronology. However, the above measure has been chosen, because \textit{quasi}-empiricism is a popularly known version of philosophy of mathematics as an alternative to philosophy confederated with the study of foundations, and a reference to it will make it relatively easy to pinpoint the ``locus'' of the other way of deliberation on mathematics -- presented by Bôcher -- on the ``map'' of variants of philosophy of mathematics.

\section{An outline of \textit{quasi}-empiricism}
The beginnings of quasi-empiricism are related to a four-part paper entitled \textit{Proofs and Refutations}
%\label{ref:RNDlCDsZHJWUQ}(Lakatos, 1963),
\parencite[][]{lakatos_proofs_1963},
 which Lakatos published in ``The British Journal of the Philosophy of Science'' in 1963-1964. A coherent description of the direction originated with Lakatos' publications is not an easy task. The reason may be that it is not so much about a uniform direction in the philosophy of mathematics, as about a wave of -- essentially uncoordinated -- reactions to the way philosophy of mathematics had been pursued before.

@@

Understandably enough, it is interesting to ask whether the ``new'' philosophy of mathematics actually had its origins only in Lakatos' above-mentioned publications in the 1960s, or was there any pre-history to it. While asking about the pre-history of \textit{quasi}-empiricism, one should first and foremost take note of the authors who inspired Lakatos. These include Hegel, with his historism and dialectic, Popper, who emphasised the fallibility of statements made by empirical sciences, and Pólya, who specialised in heuristics of mathematics. Since the philosophy of mathematics was not among the pursuits of the former two, only the Hungarian mathematician and philosopher can be reckoned among those who originated the pre-history of \textit{quasi}-empiricism. Also, Wittgenstein, as represented by the late period of his activity, and Quine are usually mentioned. Certainly, account should also be taken of the lifework of Hadamard
%\label{ref:RNDwSdgR32pBO}(1945)
\parencite*[][]{hadamard_psychology_1945}
 and Poincaré 
%\label{ref:RNDkRcc7QWJnb}(1904),
\parencite*[][]{poincare_valeur_1904},
 who pursued, inter alia, ``non-deductive,'' ``psychological'' aspects of mathematicians' ``output.''

Therefore, as already stated, even though the threads of the ``new'' philosophy were invariably present in the history of deliberation on mathematics, for more than half a century, until Lakatos' publications, they were ``muffled'' by philosophy confederated with the study of the foundations\footnote{It must be emphasised, once again, that the ``Creative Mathematician'' relying on intuition and imagination has never been out of interest for the philosophy of mathematics. This aspect of considering mathematics was represented by Husserl, and under his influence by the French school of philosophy of mathematics -- starting from Poincaré. Husserl's phenomenology (intuition) was invoked in Gödel's attempt to explain the ``perception'' of the objective infinite hierarchy of infinite sets. However, none of them stressed as strongly the fallibility of the theorems of the mathematics as Bôcher and Lakatos did. Since the thesis of the fallibility of mathematical theorems is the essence of Lakatos' \textit{quasi}-empiricism, the roots of \textit{quasi}-empiricism are to be found in Bôcher.}.

@@

\begin{enumerate}
\item The ``discussion [...] of the so-called foundations'' is a new and essential direction of reflection on mathematics.
\item However, it is extremely one-sided and inadequate for complete philosophical deliberation on mathematics;
\item That is why it needs to be complemented by another perception of mathematics, one that takes into account -- to stay in keeping with Bôcher's terminology -- (only) that which is non-deductive in it.
\end{enumerate}
With the benefit of the contextual terminology, which was introduced before, and the remark whereby logicism, formalism and intuitionism were focused on the study of the context of substantiation in mathematics, as well as the results of the analyses performed in the paper
%\label{ref:RND15HRIVaRP6}(Dadaczyński, 2015),
\parencite[][]{dadaczynski_tendencje_2015},
 one can say that what Bôcher outlines in the first part of his work -- as an approach to mathematics that is a ``discussion [...] of the so-called foundations'' -- is in fact a (one-sided) proposal to study the context of substantiation. This context can also be referred to as -- to elaborate Bôcher's terminology (the term ``non-deductive'' to be specific) –a deductive aspect of mathematics.

The alternative reflection on mathematics proposed by Bôcher is supposed to take into account -- according to his metaphilosophical declaration -- (only) a non-deductive aspect of mathematics. Further analysis in the present paper is aimed at answering the question: does the non-deductive aspect of mathematics, singled out by Bôcher, ``coincide'' with the context of discovery in mathematics?

\subsection{3.1. A ``creative mathematician'' -- significance of imagination}

Bôcher begins his characterisation of the non-deductive elements of mathematics with a comparison. He writes that he likes to perceive mathematics as more of an art than science. In his opinion, a mathematician is an artist led by, though not controlled, by the external, sensual world. Creativity thus construed bears a real resemblance to the activity engaged in by an artist, e.g. a painter. Bôcher likens the mathematician's capacity to pursue deductive reasoning to the artist's skill at mastering painting techniques. When properly developed, the said capacity is a \textit{sine qua non} for being a mathematician, just like the skill at mastering suitable painting techniques is a \textit{sine qua non} for becoming a painter. However, these capabilities are not sufficient conditions for becoming either a ``full-blooded'' mathematician or a ``full-blooded'' painter; what is more -- in Bôcher's opinion -- they are not even the most important skills conditioning becoming either of the two. Other predispositions play a crucial role, but above all imagination -- in both cases -- is a more vital factor ``making'' someone a ``good mathematician'' or a ``good painter.''\footnote{``I like to look at mathematics almost more as an art than as a science; for the activity of the mathematician, constantly creating as he is, guided though not controlled by the external world of the senses, bears a resemblance, not fanciful I believe but real, to the activity of an artist, of a painter let us say. Rigorous deductive reasoning on the part of the mathematician may be likened here to technical skill in drawing on the part of the painter. Just as no one can become a good painter without a certain amount of this skill, so no one can become a mathematician without the power to reason accurately up to a certain point. Yet these qualities, fundamental though they are, do not make a painter or a mathematician worthy of the name, nor indeed are they the most important factors in the case. Other qualities of a far more subtle sort, chief among which in both cases is imagination, go to the making of the good artist or good mathematician''
%\label{ref:RNDfe2UhwlTxs}(Bôcher, 1904, p.133).
\parencite[][p.133]{bocher_fundamental_1904}.%
}

The fact that Bôcher regards a (good) mathematician primarily as an artist is of relevance to the research in hand. And here apparently lies the fundamental similarity between Bôcher's reflection on mathematics and later reflection by many representatives of \textit{quasi}-empiricism. The latter ones were primarily interested in the ``creative'' mathematician, his process of obtaining crucial results. They tried to describe the ``creative'' mathematician and to isolate -- as far as possible -- the procedures constituting mathematical creativity.

@@

At any rate -- and this should be strongly emphasised -- given the present study indicating that Bôcher addressed essential threads characteristic of the much later \textit{quasi}-empiricism, it is vital to show that Bôcher stressed the question of a ``creative'' mathematician, which was of such great import for that direction. One might add that it was exactly addressing this issue that later on came to bear on the shift of emphasis in the philosophy of mathematics effected by representatives of \textit{quasi}-empiricism in relation to the earlier philosophy of mathematics pursued in ``the spirit of foundations.''

But it is exactly the latter statement that is potentially debatable. After all, it was within the framework of intuitionism -- as well as other constructivist trends -- that unusually dogmatic emphasis was placed on the issues concerned with creation -- constructing mathematical objects. And it was the subject who was supposed to do the constructing.

However, leaving aside the fact that there was no consensus with regard to specifying the construction, it must be concluded that within the framework of intuitionism there was no answer to the question: how does the subject (mathematician) do the constructing, and to be more precise, what are the subjective determinants of being a ``creative'' (constructing) mathematician. It was exactly these questions that were posed, and attempts were made at answering them within the framework of \textit{quasi}-empiricism and in Bôcher's text. Besides, constructivism would sometimes even ``hinder'' intuitionists by generating ``proof-theory'' restrictions. It is sufficient to point to the rejection -- for constructivist reasons -- of indirect proofs of existential theorems. A mathematician hindered by those principles certainly did not match the vision of a ``liberated'' mathematician-artist as outlined by Bôcher.

@@

Bôcher stresses that he means above all experiments in number theory (as well as in analysis). He says that experiment records -- which, as one might believe, corroborate some statements (hypotheses) in number theory -- often used to be contained in mathematicians' publications. He goes on to add that the fact that in his times the practice of publishing the records of such activities was abandoned does not change, in his opinion, the fact that an arithmetical experiment -- one might add: in the context of mathematical discovery (to be more precise: of a number-theory character), in study practice -- was as common as before.\footnote{,,Then there is the method of experiment; not merely the physical experiments of the laboratory or the geometric experiments I had occasion to speak of a few minutes ago, but also arithmetical experiments, numerous examples of which are found in the theory of numbers and in analysis. The mathematicians of the past frequently used this method in their printed works. That this is now seldom done must not be taken to indicate that the method itself is not used as much as ever''
%\label{ref:RNDQdKDMpRbg8}(Bôcher, 1904, p.134).
\parencite[][p.134]{bocher_fundamental_1904}.%
}

@@

In general, two cases are possible, and they always need to be taken into account by him who is performing the number-theory experiment: with the natural numbers substituted for the variables the expression (1) is either true, or it is not. In the first case corroboration (to use Popper's terminology) of the statement (1) takes place, while the second case is one of falsification (to use the same terminology) of the ``experimental'' -- and hence hypothetical -- statement (1).

@@

With regard to the present study, it is essential that as Bôcher fully accepts an experiment among the mathematician's instruments, above all but not only in the field of number theory investigations, and \textit{explicite} declares its relations with a physical experiment in a laboratory, he points, \textit{implicite} – but as one might surmise consciously -- to the fact that in mathematical practice some mathematical statements are accepted ``experimentally'' as hypotheses and -- as one might add today: in the context of mathematical discovery -- some elements of mathematical knowledge (some statements in which this knowledge is expressed) are fallible.

As demonstrated before, the statement whereby some mathematical propositions are fallible and as such may be subjected to testing was the main theorem of Lakatos' \textit{quasi}-empiricism, inspired by the thesis of fallibility of physical knowledge and the thesis of the actual application of falsification processes in the field of physics, the theses having been formulated by Popper. And therefore, by advancing -- though only \textit{implicite} by addressing the relevance of an experiment in mathematics (and in particular in number theory) -- the issue of fallibility of some parts of mathematical knowledge, Bôcher emphasised another crucial thesis characteristic of the later \textit{quasi}-empiricism of the 1960s.

@@

The above statements lead to two vital -- from the viewpoint of the present study -- conclusions:

>>>>

\begin{artengenv}{Jerzy Dadaczyński, Robert Piechowicz}
	{Maxime Bôcher's concept of complementary philosophy of mathematics}
	{Maxime Bôcher's concept of complementary philosophy of mathematics}
	{Maxime Bôcher's concept of complementary philosophy\\of mathematics}
	{Pontifical University of John Paul II in Krakow}
	{The main purpose of the present paper is to demonstrate that as early as 1904 pre-eminent American mathematician Maxime Bôcher was an adherent to the presently relevant argument of reasonableness, or even necessity of parallel development of two philosophical methods of reflection on mathematics, so that its essence could be more fully comprehended. The goal of the research gives rise to the question: what two types of philosophical deliberation on mathematics were proposed by Bôcher?}
	{philosophy, mathematics, Maxim Bôcher, metamathematics.}

\section{Introduction}
\lettrine[loversize=0.13,lines=2,lraise=-0.05,nindent=0em,findent=0.2pt]%
{T}{}he main purpose of the present paper is to demonstrate that as early as 1904 pre-eminent American mathematician Maxime Bôcher\footnote{Maxime Bôcher (1867-1918) was a son of an American scientist of French origin. In the years 1883-1888 he studied, \textit{inter alia}, mathematics at Harvard. In the years 1888-1891, in Göttingen he was a student of Klein's, under the direction of whom he wrote and defended his doctoral dissertation. After his return to the USA he was an assistant professor, and as of 1904 a professor of mathematics at Harvard. In the years 1909-1910 he served as president of the American Mathematical Society. Bôcher specialised chiefly in mathematical analysis, and his most famous paper is \textit{Introduction to Higher Algebra}
%\label{ref:RNDc7ZxUmWWTi}(Bôcher, 1907).
\parencite[][]{bocher_introduction_1907}.
To honour the memory of this mathematician, who died prematurely, Bôcher Memorial Prize was endowed; it is awarded at intervals of several years for achievements in the field of mathematical analysis in the period of the six years leading up to the prize-giving year. One of the prize-winners was von Neumann (in the 1930s).} was an adherent to the presently relevant argument of reasonableness, or even necessity of parallel development of two philosophical---complementary---methods of reflection on mathematics, so that its essence could be more fully comprehended.

Thus outlined, the main goal of the research gives rise to the question: what two---complementary---types of philosophical deliberation on mathematics were proposed by Bôcher? They were ``adumbrated'' in the text entitled \textit{The Fundamental Conceptions and Methods of Mathematics}
%\label{ref:RNDNx8Z5TUKyi}(Bôcher, 1904).
\parencite[][]{bocher_fundamental_1904}.

In the first, principal part of his text---which has already been analysed in

@@

\parencite[][]{dadaczynski_tendencje_2015}---Bôcher made a direct reference to logicism, proposing its corrections, so that the reductionist programme could be appropriately realised.\footnote{Bôcher was doubtful as to the possibility of realising the programme of logicism the way it had been delineated by Frege and Russell. His objection was that logic---as viewed by the originators of logicism---was too ``weak'' to ``produce'' an infinite set of natural numbers. Bôcher's predictions proved right while work on \textit{Principia Mathematica} was underway: it was necessary to ``add'' some form of an axiom of infinity---in the ontological version providing for the existence of an infinitude of individuals---to obtain natural numbers (of any type). Such an axiom did not ``fit within'' Frege's logic or Russell's ``starting'' logic. The American mathematician formulated his objections as follows: ``Nevertheless, in view of the fact that the system of finite positive integers is necessary in almost all branches of mathematics (we cannot speak of a triangle or a hexagon without having the numbers three and six at our disposal), it seems extremely desirable that the system of logical principles which we lay at the foundation of all mathematics be assumed, if possible, broad enough so that the existence of positive integers---at least finite integers---follows from it; and there seems little doubt that this can be done in a satisfactory manner. When this has been done we shall perhaps be able to regard, with Russell, pure mathematics as consisting exclusively of deductions by ‘logical principles from logical principles'' 

@@

} He made an \textit{explicite} demand for future demonstration of non-contradiction of all mathematics liberated---and such liberation he anticipated---of known antinomies, and not just ``partial'' demands for proof of non-contradiction of the arithmetic of real numbers 

@@

In a sense---which was demonstrated in the paper 

@@

\parencite[][]{dadaczynski_tendencje_2015}---the first part of Bôcher's text contains, \textit{inter alia}, an ``outline,'' though an unordered one, of the 1904 state of philosophy of mathematics confederated with the study of its foundations.\footnote{The passage below serves to bring back the conclusions from the paper 

@@

 In 1904 logicism was already a well-developed direction in the study of the foundations of mathematics. Bôcher could see its intrinsic difficulties concerned with the establishment of antinomies, but he was sure that corrections of the logic underlying the foundations of mathematics would make it possible to remove those difficulties. Hence, in a sense, he ``announced'' the ``second'' logicism---the one of Russell's. Two points excerpted from the 1904 text foreshadowed Hilbert's formalism of the 1920s. They are, firstly, methodical nominalism, and, secondly, posing a question whether it is possible to establish---having removed known antinomies---a global non-contradiction of mathematics. The latter issue was Bôcher's clear reference to Hilbert's second problem (non-contradiction of the arithmetic of real numbers) and Hilbert's first attempts at proving the non-contradiction of the arithmetic of natural numbers. The positive solution to the problem of the global non-contradiction of mathematics---which was emphasised by, \textit{inter alia}, Bôcher---later on became the goal of the formalist research programme, the attainment of which being the rationale for the emergence of metamathematics. Methodological nominalism became Hilbert's answer to the question about semantics of the fully axiomatised and formalised classical mathematics. Bôcher's text of 1904 negatively treats Kantianism as the foundation on which to construct mathematics. Nevertheless, it is a source that confirms that Kantianism was still relevant in many mathematical milieus in the early 20\textsuperscript{th} century. That was the ground on which, several years after the publication of Bôcher's text and following the adoption of appropriate corrections and modifications, intuitionism emerged.} Noteworthily, Bôcher termed his own way of deliberating over mathematics, which he presented in the first, principal part of his work, ``discussion [...] of the so-called foundations'' 

@@

 which focuses on the deductive aspect of mathematics. His---to use his own term---``discussion [...] of the so-called foundations'' of mathematics, in which he develops some of his own threads, \textit{inter alia}, attempting to specify what mathematics is, is a kind of reflection markedly determined by the way (and ``contents'') of Frege's, Russell's and Hilbert's deliberation on mathematics.

Since the first way of deliberation on mathematics, which was proposed by Bôcher, has already been subjected to analysis and identified, demonstrating that his metaphilosophy contained a demand for inclusion of the two---complementary---types of philosophy of mathematics becomes reduced to showing that Bôcher was outlining an alternative---though naturally not in the sense of an exclusive alternative---concept of philosophical reflection on mathematics, which in his opinion was to complement philosophy confederated with the study of the foundations of mathematics.

With a view to the attainment of the above-indicated research objective, in the first place, presentation will be made of the main outlines of \textit{quasi}-empiricism---philosophy of mathematics initiated with Lakatos' publications of the first half of the 1960s. The fundamental imperatives of \textit{quasi}-empiricism shall serve here as points of reference for analysis and identification of the content of the alternative variant of philosophy of mathematics outlined by Bôcher.

This place calls for a crucial remark. Originated with Lakatos' texts, \textit{quasi}-empiricism strongly emphasised some threads of philosophy of mathematics which, although constantly present in deliberation on mathematics, for more than half a century were ``suppressed'' by study concerned with foundations (logicism, formalism and intuitionism). Mention will be made here of the prehistory of \textit{quasi}-empiricism, and one of partial propositions will be to show that Bôcher should be reckoned among its representatives. The choice of \textit{quasi}-empiricism as a point of reference for the study of an alternative thread of Bôcher's philosophy of mathematics may seem to be an ahistorical approach exactly because it does not take account of the above-mentioned chronology. However, the above measure has been chosen, because \textit{quasi}-empiricism is a popularly known version of philosophy of mathematics as an alternative to philosophy confederated with the study of foundations, and a reference to it will make it relatively easy to pinpoint the ``locus'' of the other way of deliberation on mathematics---presented by Bôcher---on the ``map'' of variants of philosophy of mathematics.

\section{An outline of \textit{quasi}-empiricism}
The beginnings of quasi-empiricism are related to a four-part paper entitled \textit{Proofs and Refutations}
%\label{ref:RNDlCDsZHJWUQ}(Lakatos, 1963),
\parencite[][]{lakatos_proofs_1963},
 which Lakatos published in ``The British Journal of the Philosophy of Science'' in 1963-1964. A coherent description of the direction originated with Lakatos' publications is not an easy task. The reason may be that it is not so much about a uniform direction in the philosophy of mathematics, as about a wave of---essentially uncoordinated---reactions to the way philosophy of mathematics had been pursued before.

@@

Understandably enough, it is interesting to ask whether the ``new'' philosophy of mathematics actually had its origins only in Lakatos' above-mentioned publications in the 1960s, or was there any pre-history to it. While asking about the pre-history of \textit{quasi}-empiricism, one should first and foremost take note of the authors who inspired Lakatos. These include Hegel, with his historism and dialectic, Popper, who emphasised the fallibility of statements made by empirical sciences, and Pólya, who specialised in heuristics of mathematics. Since the philosophy of mathematics was not among the pursuits of the former two, only the Hungarian mathematician and philosopher can be reckoned among those who originated the pre-history of \textit{quasi}-empiricism. Also, Wittgenstein, as represented by the late period of his activity, and Quine are usually mentioned. Certainly, account should also be taken of the lifework of Hadamard
%\label{ref:RNDwSdgR32pBO}(1945)
\parencite*[][]{hadamard_psychology_1945}
 and Poincaré 
%\label{ref:RNDkRcc7QWJnb}(1904),
\parencite*[][]{poincare_valeur_1904},
 who pursued, \textit{inter alia}, ``non-deductive,'' ``psychological'' aspects of mathematicians' ``output.''

Therefore, as already stated, even though the threads of the ``new'' philosophy were invariably present in the history of deliberation on mathematics, for more than half a century, until Lakatos' publications, they were ``muffled'' by philosophy confederated with the study of the foundations\footnote{It must be emphasised, once again, that the ``Creative Mathematician'' relying on intuition and imagination has never been out of interest for the philosophy of mathematics. This aspect of considering mathematics was represented by Husserl, and under his influence by the French school of philosophy of mathematics---starting from Poincaré. Husserl's phenomenology (intuition) was invoked in Gödel's attempt to explain the ``perception'' of the objective infinite hierarchy of infinite sets. However, none of them stressed as strongly the fallibility of the theorems of the mathematics as Bôcher and Lakatos did. Since the thesis of the fallibility of mathematical theorems is the essence of Lakatos' \textit{quasi}-empiricism, the roots of \textit{quasi}-empiricism are to be found in Bôcher.}.

@@

\begin{enumerate}
\item The ``discussion [...] of the so-called foundations'' is a new and essential direction of reflection on mathematics.
\item However, it is extremely one-sided and inadequate for complete philosophical deliberation on mathematics;
\item That is why it needs to be complemented by another perception of mathematics, one that takes into account---to stay in keeping with Bôcher's terminology---(only) that which is non-deductive in it.
\end{enumerate}
With the benefit of the contextual terminology, which was introduced before, and the remark whereby logicism, formalism and intuitionism were focused on the study of the context of substantiation in mathematics, as well as the results of the analyses performed in the paper
%\label{ref:RND15HRIVaRP6}(Dadaczyński, 2015),
\parencite[][]{dadaczynski_tendencje_2015},
 one can say that what Bôcher outlines in the first part of his work---as an approach to mathematics that is a ``discussion [...] of the so-called foundations''---is in fact a (one-sided) proposal to study the context of substantiation. This context can also be referred to as---to elaborate Bôcher's terminology (the term ``non-deductive'' to be specific) –a deductive aspect of mathematics.

The alternative reflection on mathematics proposed by Bôcher is supposed to take into account---according to his metaphilosophical declaration---(only) a non-deductive aspect of mathematics. Further analysis in the present paper is aimed at answering the question: does the non-deductive aspect of mathematics, singled out by Bôcher, ``coincide'' with the context of discovery in mathematics?

\subsection{3.1. A ``creative mathematician''---significance of imagination}

Bôcher begins his characterisation of the non-deductive elements of mathematics with a comparison. He writes that he likes to perceive mathematics as more of an art than science. In his opinion, a mathematician is an artist led by, though not controlled, by the external, sensual world. Creativity thus construed bears a real resemblance to the activity engaged in by an artist, e.g. a painter. Bôcher likens the mathematician's capacity to pursue deductive reasoning to the artist's skill at mastering painting techniques. When properly developed, the said capacity is a \textit{sine qua non} for being a mathematician, just like the skill at mastering suitable painting techniques is a \textit{sine qua non} for becoming a painter. However, these capabilities are not sufficient conditions for becoming either a ``full-blooded'' mathematician or a ``full-blooded'' painter; what is more---in Bôcher's opinion---they are not even the most important skills conditioning becoming either of the two. Other predispositions play a crucial role, but above all imagination---in both cases---is a more vital factor ``making'' someone a ``good mathematician'' or a ``good painter.''\footnote{``I like to look at mathematics almost more as an art than as a science; for the activity of the mathematician, constantly creating as he is, guided though not controlled by the external world of the senses, bears a resemblance, not fanciful I believe but real, to the activity of an artist, of a painter let us say. Rigorous deductive reasoning on the part of the mathematician may be likened here to technical skill in drawing on the part of the painter. Just as no one can become a good painter without a certain amount of this skill, so no one can become a mathematician without the power to reason accurately up to a certain point. Yet these qualities, fundamental though they are, do not make a painter or a mathematician worthy of the name, nor indeed are they the most important factors in the case. Other qualities of a far more subtle sort, chief among which in both cases is imagination, go to the making of the good artist or good mathematician''
%\label{ref:RNDfe2UhwlTxs}(Bôcher, 1904, p.133).
\parencite[][p.133]{bocher_fundamental_1904}.%
}

The fact that Bôcher regards a (good) mathematician primarily as an artist is of relevance to the research in hand. And here apparently lies the fundamental similarity between Bôcher's reflection on mathematics and later reflection by many representatives of \textit{quasi}-empiricism. The latter ones were primarily interested in the ``creative'' mathematician, his process of obtaining crucial results. They tried to describe the ``creative'' mathematician and to isolate---as far as possible---the procedures constituting mathematical creativity.

@@

At any rate---and this should be strongly emphasised---given the present study indicating that Bôcher addressed essential threads characteristic of the much later \textit{quasi}-empiricism, it is vital to show that Bôcher stressed the question of a ``creative'' mathematician, which was of such great import for that direction. One might add that it was exactly addressing this issue that later on came to bear on the shift of emphasis in the philosophy of mathematics effected by representatives of \textit{quasi}-empiricism in relation to the earlier philosophy of mathematics pursued in ``the spirit of foundations.''

But it is exactly the latter statement that is potentially debatable. After all, it was within the framework of intuitionism---as well as other constructivist trends---that unusually dogmatic emphasis was placed on the issues concerned with creation---constructing mathematical objects. And it was the subject who was supposed to do the constructing.

However, leaving aside the fact that there was no consensus with regard to specifying the construction, it must be concluded that within the framework of intuitionism there was no answer to the question: how does the subject (mathematician) do the constructing, and to be more precise, what are the subjective determinants of being a ``creative'' (constructing) mathematician. It was exactly these questions that were posed, and attempts were made at answering them within the framework of \textit{quasi}-empiricism and in Bôcher's text. Besides, constructivism would sometimes even ``hinder'' intuitionists by generating ``proof-theory'' restrictions. It is sufficient to point to the rejection---for constructivist reasons---of indirect proofs of existential theorems. A mathematician hindered by those principles certainly did not match the vision of a ``liberated'' mathematician-artist as outlined by Bôcher.

@@

Bôcher stresses that he means above all experiments in number theory (as well as in analysis). He says that experiment records---which, as one might believe, corroborate some statements (hypotheses) in number theory---often used to be contained in mathematicians' publications. He goes on to add that the fact that in his times the practice of publishing the records of such activities was abandoned does not change, in his opinion, the fact that an arithmetical experiment---one might add: in the context of mathematical discovery (to be more precise: of a number-theory character), in study practice---was as common as before.\footnote{,,Then there is the method of experiment; not merely the physical experiments of the laboratory or the geometric experiments I had occasion to speak of a few minutes ago, but also arithmetical experiments, numerous examples of which are found in the theory of numbers and in analysis. The mathematicians of the past frequently used this method in their printed works. That this is now seldom done must not be taken to indicate that the method itself is not used as much as ever''
%\label{ref:RNDQdKDMpRbg8}(Bôcher, 1904, p.134).
\parencite[][p.134]{bocher_fundamental_1904}.%
}

@@

In general, two cases are possible, and they always need to be taken into account by him who is performing the number-theory experiment: with the natural numbers substituted for the variables the expression (1) is either true, or it is not. In the first case corroboration (to use Popper's terminology) of the statement (1) takes place, while the second case is one of falsification (to use the same terminology) of the ``experimental''---and hence hypothetical---statement (1).

@@

With regard to the present study, it is essential that as Bôcher fully accepts an experiment among the mathematician's instruments, above all but not only in the field of number theory investigations, and \textit{explicite} declares its relations with a physical experiment in a laboratory, he points, \textit{implicite} – but as one might surmise consciously---to the fact that in mathematical practice some mathematical statements are accepted ``experimentally'' as hypotheses and---as one might add today: in the context of mathematical discovery---some elements of mathematical knowledge (some statements in which this knowledge is expressed) are fallible.

As demonstrated before, the statement whereby some mathematical propositions are fallible and as such may be subjected to testing was the main theorem of Lakatos' \textit{quasi}-empiricism, inspired by the thesis of fallibility of physical knowledge and the thesis of the actual application of falsification processes in the field of physics, the theses having been formulated by Popper. And therefore, by advancing---though only \textit{implicite} by addressing the relevance of an experiment in mathematics (and in particular in number theory)---the issue of fallibility of some parts of mathematical knowledge, Bôcher emphasised another crucial thesis characteristic of the later \textit{quasi}-empiricism of the 1960s.

@@

The above statements lead to two vital---from the viewpoint of the present study---conclusions:
